The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-08-02 Independence_CH/details.tex
Sit-rep on manuscript. Eliminating original comments by Larry and repeated arity thms

\subsection{36 replacement instances to rule them all}
\label{sec:repl-instances}

We isolated 36 instances of Replacement that are sufficient to force
$\CH$ or $\neg\CH$. The first 15 instances are needed to set up
cardinal arithmetic in $M$. Many of these were already present in
relational form in the \session{ZF-Constructible} library.

\begin{itemize}
\item 8 instances for iteration of operations. For each construct,
  Paulson used one replacement instance in order to have absoluteness,
  and a second one to obtain closure.
  \begin{itemize}
  \item
    Lists
    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.
  \item
    Formulas
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.
  \item
    Transitive closure %(iteration of $\union$)
    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
  \item
    $n$th element of a list (iteration of the tail operation)
    \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}. By transitivity,
    we do not need an instance for closure; the only instance was required
    for absoluteness.
  \item Absoluteness of the recursive construction in the proof of the
    Schröder-Bernstein Theorem \isatt{banach{\uscore}iterates{\uscore}fm}.
  \end{itemize}
\item
  2 instances for definition of ordertypes.
  \begin{itemize}
  \item \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
  \item Replacement through $x\mapsto \otype(x)$ \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
  \end{itemize}
\item 1 instance for definition of cardinal successor.
  \begin{itemize}
  \item
    \isatt{replacement{\uscore}is{\uscore}jump{\uscore}cardinal{\uscore}body{\uscore}fm}.
  \end{itemize}
\item
  4 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item Ordertypes
    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
  \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
  \item Cumulative hierarchy (rank initial segments) \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
  \item Aleph \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
  \end{itemize}
\end{itemize}

@@

That makes 30 instances up to now. For the setup of forcing, we
require the following 5 instances (the first two are needed for the $\Delta$-System Lemma):

\begin{itemize}
%\verb|replacement_ax2(6)| \verb|replacement_transrec_apply_image_body_fm| seems fake!
\item Recursive construction of sets using a choice function
  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
% \item Replacement through  $x\mapsto \check{x}$ \isatt{check{\uscore}replacement{\uscore}fm}.
%\item $x\mapsto |x|^M$
%  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(is{\uscore}cardinal{\uscore}fm(0,1),0)}. % fake since the image is known
\item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
\item 2 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item
    Check names
    \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
  \item Forcing for atomic formulas
    \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
  \end{itemize}
\item Replacement through $x\mapsto \lb\check{x},x\rb$ (for the
  definition of $\punto{G}$)
  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
\end{itemize}
%
This is enough to force $\neg\CH$. To force $\CH$, one further instance is needed:
%
\begin{itemize}
\item Absoluteness of the recursive construction in the proof of the
  Dependent Choices \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
\end{itemize}

The need for some of the instances above arose from the particular
engineering of the some of the proofs. For instance, the construction
of the cardinal successor should go through by reusing the replacement
$x\mapsto \otype(x)$. Nevertheless, we relativized Paulson's
approach direclty, which had an alternative formulation to obtain a
simpler proof of Hartogs' Theorem (in particular, that $\kappa^+$ is
indeed a cardinal).
% https://isabelle.in.tum.de/dist/library/ZF/ZF/CardinalArith.html#CardinalArith.jump_cardinal_def|fact

>>>>

\subsection{34 replacement instances to rule them all}
\label{sec:repl-instances}

We isolated 34 instances of Replacement that are sufficient to force
$\CH$ or $\neg\CH$. The first 14 instances are needed to set up
cardinal arithmetic in $M$. Many of these were already present in
relational form in the \session{ZF-Constructible} library.

\begin{itemize}
\item 6 instances for iteration of operations. For each construct,
  Paulson used one replacement instance in order to have absoluteness,
  and a second one to obtain closure.
  \begin{itemize}
  \item
    Lists
    \isatt{list{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{list{\uscore}repl2{\uscore}intf{\uscore}fm}.
  \item
    Formulas
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{formula{\uscore}repl1{\uscore}intf{\uscore}fm}.
  \item
    Transitive closure %(iteration of $\union$)
    \isatt{eclose{\uscore}repl1{\uscore}intf{\uscore}fm},
    \isatt{eclose{\uscore}repl2{\uscore}intf{\uscore}fm}.
  \end{itemize}
\item 2 instances for absoluteness of iterative constructions.
  \begin{itemize}
  \item
    $n$th element of a list (iteration of the tail operation)
    \isatt{tl{\uscore}repl{\uscore}intf{\uscore}fm}. By transitivity,
    we do not need an instance for closure.
  \item Absoluteness of the recursive construction in the proof of the
    Schröder-Bernstein Theorem \isatt{banach{\uscore}iterates{\uscore}fm}.
  \end{itemize}
\item
  2 instances for ordertypes.
  \begin{itemize}
  \item Auxiliary for the definition of ordertypes
    \isatt{replacement{\uscore}is{\uscore}order{\uscore}eq{\uscore}map{\uscore}fm}.
  \item Replacement through $x\mapsto \otype(x)$, for Hartogs' Theorem
    \isatt{replacement{\uscore}is{\uscore}order{\uscore}body{\uscore}fm}.
  \end{itemize}
\item
  4 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item Ordertypes
    \isatt{wfrec{\uscore}replacement{\uscore}order{\uscore}pred{\uscore}fm}.
  \item Rank \isatt{wfrec{\uscore}rank{\uscore}fm}.
  \item Cumulative hierarchy (rank initial segments) \isatt{trans{\uscore}repl{\uscore}HVFrom{\uscore}fm}.
  \item Aleph \isatt{replacement{\uscore}HAleph{\uscore}wfrec{\uscore}repl{\uscore}body{\uscore}fm}.
  \end{itemize}
\end{itemize}

@@

That makes 28 instances up to now. For the setup of forcing, we
require the following 5 instances (the last two are needed for the $\Delta$-System Lemma):

\begin{itemize}
\item 2 instances for definitions by well-founded recursion.
  \begin{itemize}
  \item
    Check names
    \isatt{wfrec{\uscore}Hcheck{\uscore}fm}.
  \item Forcing for atomic formulas
    \isatt{wfrec{\uscore}Hfrc{\uscore}at{\uscore}fm}.
  \end{itemize}
\item Replacement through $x\mapsto \lb x,\check{x}\rb$ (for the
  definition of $\punto{G}$)
  \isatt{Lambda{\uscore}in{\uscore}M{\uscore}fm(check{\uscore}fm(2,0,1),1)}.
\item Recursive construction of sets using a choice function
  \isatt{replacement{\uscore}is{\uscore}trans{\uscore}apply{\uscore}image{\uscore}fm}.
\item Absoluteness of the previous construction \isatt{replacement{\uscore}transrec{\uscore}apply{\uscore}image{\uscore}body{\uscore}fm}.
\end{itemize}
%
This is enough to force $\neg\CH$. To force $\CH$, one further instance is needed:
%
\begin{itemize}
\item Absoluteness of the recursive construction in the proof of the
  Dependent Choices \isatt{replacement{\uscore}dcwit{\uscore}repl{\uscore}body{\uscore}fm}.
\end{itemize}

The need for some of the instances above arose from the particular
engineering of the some of the proofs. For instance, the first
instance required for the definition of relative ordertypes arises
from Paulson's \session{ZF-Constructible}.
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Rank.html#offset_1123..1139
